---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
library(data.table)
library(ggplot2)
library(knitr)
```

## Question 1

```{r}
pw_pca <- function (data){
  ## function to do principle component analysis
  ## param data: data on which to do PCA
  ## returns: list with eigenvectors, correlation coefficients, variance and cumulative variances
  
  p <- ncol(data)
  sigma <- cov(data)
  xbar <- colMeans(data)
  
  eigen_vectors <- eigen(sigma)$vectors
  eigen_values <- eigen(sigma)$values
  cum_var <- matrix(,ncol=1,nrow=p)
  cum_var[1,] <- (eigen_values[1]/sum(eigen_values))
  
  for (i in 2:p){
    cum_var[i,] <- (eigen_values[i]/sum(eigen_values))+cum_var[i-1]
  }
  
  n <- sum(cum_var<0.97)
  rho <- matrix(,ncol=n,nrow=p)
  
  for (i in 1:p){
    for (j in 1:n){
      rho[i,]<-(eigen_vectors[i,j]*sqrt(eigen_values[j]))/sqrt(sigma[i,i])
    }
  }
  
  par(mfrow=c(2,1))
  
  y1<-(eigen_vectors[1,1]*(data[,1]-xbar[1])) + (eigen_vectors[2,1]*(data[,2]-xbar[2]))+(eigen_vectors[3,1]*(data[,3]-xbar[3]))+(eigen_vectors[4,1]*(data[,4]-xbar[4]))+(eigen_vectors[5,1]*(data[,5]-xbar[5]))
  y2<-(eigen_vectors[1,2]*(data[,1]-xbar[1])) + (eigen_vectors[2,2]*(data[,2]-xbar[2]))+(eigen_vectors[3,2]*(data[,3]-xbar[3]))+(eigen_vectors[4,2]*(data[,4]-xbar[4]))+(eigen_vectors[5,2]*(data[,5]-xbar[5]))
  
  
  pca_plot <- ggplot() + geom_point(aes(x=y2$V1,y=y1$V1)) + ggtitle("Plot of first two principal components") + ylab("yhat 2") + xlab("yhat 1")
  print(pca_plot)
  
  scree_plot <- ggplot() + geom_point(aes(y=eigen_values, x=seq(1,length(eigen_values))))+ geom_line(aes(y=eigen_values, x=seq(1,length(eigen_values)))) + xlab("i") + ggtitle("Scree plot") + ylab("eigen values")
  print(scree_plot)
  
  return(list("eigen vectors"=eigen_vectors,"correlation coefficients"=rho, "variance"=eigen_values,"cumulative variances"=cum_var))
}

table_8_5 <- fread('T8-5.dat')

pw_pca(table_8_5)
```

```{r}

pw_pca_2 <- function (data){
  ## function to do principle component analysis
  ## param data: data on which to do PCA
  ## returns: list with eigenvectors, correlation coefficients, variance and cumulative variances
  
  p <- ncol(data)
  sigma <- cov(data)
  xbar <- colMeans(data)
  
  eigen_vectors <- eigen(sigma)$vectors
  eigen_values <- eigen(sigma)$values
  cum_var <- matrix(,ncol=1,nrow=p)
  cum_var[1,] <- (eigen_values[1]/sum(eigen_values))
  
  for (i in 2:p){
    cum_var[i,] <- (eigen_values[i]/sum(eigen_values))+cum_var[i-1]
  }
  
  n <- sum(cum_var<0.97)
  rho <- matrix(,ncol=n,nrow=p)
  
  for (i in 1:p){
    for (j in 1:n){
      rho[i,]<-(eigen_vectors[i,j]*sqrt(eigen_values[j]))/sqrt(sigma[i,i])
    }
  }
  
  par(mfrow=c(2,1))
  
  y1<-(eigen_vectors[1,1]*(data[,1]-xbar[1])) + (eigen_vectors[2,1]*(data[,2]-xbar[2]))+(eigen_vectors[3,1]*(data[,3]-xbar[3]))
  y2<-(eigen_vectors[1,2]*(data[,1]-xbar[1])) + (eigen_vectors[2,2]*(data[,2]-xbar[2]))+(eigen_vectors[3,2]*(data[,3]-xbar[3]))
  
  pca_plot <- ggplot() + geom_point(aes(x=y2$V1,y=y1$V1)) + ggtitle("Plot of first two principal components") + ylab("yhat 2") + xlab("yhat 1")
  print(pca_plot)
  
  scree_plot <- ggplot() + geom_point(aes(y=eigen_values, x=seq(1,length(eigen_values))))+ geom_line(aes(y=eigen_values, x=seq(1,length(eigen_values)))) + xlab("i") + ggtitle("Scree plot") + ylab("eigen values")
  print(scree_plot)
  
  return(list("eigen vectors"=eigen_vectors,"correlation coefficients"=rho, "variance"=eigen_values,"cumulative variances"=cum_var))
}

table_6_9 <- fread('T6-9.dat')
table_6_9 <- table_6_9[,c(1:3)]

pw_pca_2(log(table_6_9))
```
## Example 8.3
```{r}
ex_8_3<-princomp(table_8_5,cor=F,scores=T)
summary(ex_8_3)
```
```{r}
pca_plot_8_3 <- ggplot() + geom_line(aes(y=ex_8_3$sdev^2,x=seq(1,length(ex_8_3$sdev)))) + geom_point(aes(y=ex_8_3$sdev^2,x=seq(1,length(ex_8_3$sdev))))
print(pca_plot_8_3)
```
```{r}
loadings(ex_8_3)
```
These results are the same as before. The scores and eigenvectors have different signs. Some of the loadings are not displayed by R due to being very small.

## Example 8.4
```{r}
ex_8_4<-princomp(table_6_9,cor=F,scores=T)
summary(ex_8_4)
```
```{r}
pca_plot_8_4 <- ggplot() + geom_line(aes(y=ex_8_4$sdev^2,x=seq(1,length(ex_8_4$sdev)))) + geom_point(aes(y=ex_8_4$sdev^2,x=seq(1,length(ex_8_4$sdev))))
print(pca_plot_8_4)
```
```{r}
loadings(ex_8_4)
```
These results are the same as before. The scores and eigenvectors have different signs.

## Question 3

```{r}

example_8_5 <- function (data){
  p <- ncol(data)
  n <- nrow(data)
  sigma <- cov(data)
  xbar <- colMeans(data)
  yhat <- matrix(,ncol=p,nrow=n)
  
  eigen_vectors <- eigen(sigma)$vectors
  eigen_values <- eigen(sigma)$values
  
  for(i in 1:p){
    yhat[,i] <- (eigen_vectors[1,i]*(data[,1]-xbar[1])) + (eigen_vectors[2,i]*(data[,2]-xbar[2]))+(eigen_vectors[3,i]*(data[,3]-xbar[3]))+(eigen_vectors[4,i]*(data[,4]-xbar[4]))+(eigen_vectors[5,i]*(data[,5]-xbar[5]))
  }
  
  data_subset <- yhat[,1:2]
  x_bar <- as.matrix(colMeans(data_subset))
  sinv <- solve(var(data_subset))
  Q <- qchisq(0.05,2,lower.tail=FALSE)
  xmin <- -5000;xmax<-5000
  ymin <- -5000;ymax<-5000
  x1 <- seq(xmin,xmax,length=100)
  x2 <- seq(ymin,ymax,length=100)
  grid_data <- as.matrix(expand.grid(x1,x2))
  ng <- nrow(grid.data)
  np <- length(x1)
  signgrid <- matrix(rep(0,ng),nrow=ng,ncol=1)
  
  for (i in 1:ng){
    signgrid[i,] <- sign((t(grid_data[i,]-x_bar)%*%sinv%*%(grid_data[i,]-x_bar))-Q) 
  }
  
  plot(data_subset[,1],data_subset[,2],xlim=c(xmin,xmax),ylim=c(ymin,ymax),xlab="yhat1",ylab="yhat2",col="red",pch=16,cex=1.25)
  
  par(new=T)
  contour(x1,x2,matrix(signgrid,np),add=TRUE,drawlabels=FALSE,levels=0,lty=2,lwd=2,col="black")
  
  mat<-cbind(signgrid,grid_data)
  
  region<-matrix(mat[mat[,1]==-1],ncol=3)
  par(new=T)
  plot(region[,2:3],xlim=c(xmin,xmax),ylim=c(ymin,ymax),col="red",xlab="yhat1",ylab="yhat2",pch=".",main="Ellipse" )
  
  list("eigen vectors"=eigen_vectors,"variance"=eigen_values,"y hat"=yhat)
}

table_5_8 <- fread('T5-8.dat')

example_8_5(table_5_8)
```
The results are slightly different to the textbook. This is because the signs of the second eigenvector are different to the textbook. Therefore, the $\hat{y}_2 values also have the opposite sign, which causes the plot to be slightly different. The same point is still identified as before.

## Question 4

```{r}

example_8_11 <- function (data){
  p<-ncol(data)
  n<-nrow(data)
  sigma<-cov(data)
  eig.vecs<-eigen(sigma)$vectors
  vars<-eigen(sigma)$values
  xbar<-apply(data,2,mean)
  yhat<-matrix(,ncol=p,nrow=n)
  for ( i in 1:p){
    yhat[,i]<-(eig.vecs[1,i]*(data[,1]-xbar[1])) + (eig.vecs[2,i]*(data[,2]-xbar[2]))+(eig.vecs[3,i]*(data[,3]-xbar[3]))+(eig.vecs[4,i]*(data[,4]-xbar[4]))+(eig.vecs[5,i]*(data[,5]-xbar[5]))
  }
  vals<-(yhat[,3]^2/vars[3])+(yhat[,4]^2/vars[4])+(yhat[,5]^2/vars[5])
  plot(vals,xlab="Period",ylab="T^2",ylim=c(0,10))
  lines(vals,col="black")
  Ch1<-qchisq(0.05,3,lower=FALSE)
  abline(h=Ch1,lty=2)
  abline(h=0,lty=3)
}


example_8_11(table_5_8)
```
## Question 5
```{r}

s<-matrix(c(5,2,2,2),byrow = T,ncol=2)

eig_vec<-eigen(s)$vectors

var<-eigen(s)$values

cum_var<-(var[1]/sum(var))

s<-matrix(c(5,2,2,2),byrow = T,ncol=2)

rho<-cov2cor(s)

eigen(rho)$vector

eigen_val<-eigen(rho)$values

eigen_val[1]/(eigen_val[1]+eigen_val[2])
```
They are not the same. The eigenvalues and vectors calculated from the correlation and covariance matrices should be different. This emphasizes that the principle components are not simple functions of one another.

```{r}

eigen_vec<-eigen(pho)$vector

eigen_vec[1,1]*sqrt(eigen_val[1])

eigen_vec[2,1]*sqrt(eigen_val[1])

eigen_vec[1,2]*sqrt(eigen_val[2])
```
```{r}

data <- matrix(c(108.28, 17.05, 1484.10,152.36, 16.59,  750.33,95.04, 10.91,  766.42,65.45, 14.14, 1110.46,62.97,  9.52, 1031.29,263.99, 25.33,  195.26,265.19, 18.54,  193.83,285.06, 15.73,  191.11,92.01,  8.10, 1175.16,165.68, 11.13,  211.15), byrow = TRUE, ncol = 3)

example_4_12 <- function (data){
  p<-ncol(data)
  n<-nrow(data)
  sigma<-cov(data)
  eig.vecs<-eigen(sigma)$vectors
  vars<-eigen(sigma)$values
  xbar<-apply(data,2,mean)
  cum.var<-matrix(,ncol=1,nrow=p)
  cum.var[1,]<-(vars[1]/sum(vars))
  for (i in 2:p){
    cum.var[i,]<-(vars[i]/sum(vars))+cum.var[i-1]
  }
  yhat<-matrix(,ncol=p,nrow=n)
  for ( i in 1:p){
    yhat[,i]<-(eig.vecs[1,i]*(data[,1]-xbar[1])) + (eig.vecs[2,i]*(data[,2]-xbar[2]))
  }
  yhat <<- yhat
  #Ellipse
  data.use<-yhat[,1:2]
  sinv<-solve(var(data))
  Q<-1.4
  xmin<-0;xmax<-max(data[,1])
  ymin<-min(data[,2]);ymax<-max(data[,2])
  x1<-seq(xmin,xmax,length=100)
  x2<-seq(ymin,ymax,length=100)
  grid.data<-as.matrix(expand.grid(x1,x2))
  ng<-nrow(grid.data)
  np<-length(x1)
  signgrid<-matrix(rep(0,ng),nrow=ng,ncol=1)
  for (i in 1:ng){
    signgrid[i,]<-sign((t(grid.data[i,]-xbar)%*%sinv%*%(grid.data[i,]-xbar))-Q) 
  }
  plot(data[,1],data[,2],xlim=c(xmin,xmax),ylim=c(ymin,ymax),xlab="x1",ylab="x2",
       col="red",pch=16,cex=1.25)
  par(new=T)
  contour(x1,x2,matrix(signgrid,np),add=TRUE,drawlabels=FALSE,levels=0,lty=2,lwd=2,
          col="black")
  mat<-cbind(signgrid,grid.data)
  region<-matrix(mat[mat[,1]==-1],ncol=3)
  par(new=T)
  plot(region[,2:3],xlim=c(xmin,xmax),ylim=c(ymin,ymax),col="red",xlab="x1",
       ylab="x2",pch=".",main="Ellipse" )	 
  c<-sqrt(1.4)
  one<-c*sqrt(vars[1])*eig.vecs[,1]
  two<-c*sqrt(vars[2])*eig.vecs[,2]
  lines(c(xbar[1]-one[1],xbar[1]+one[1]),c(xbar[2]-one[2],xbar[2]+one[2]),lty=1)
  lines(c(xbar[1]-two[1],xbar[1]+two[1]),c(xbar[2]-two[2],xbar[2]+two[2]),lty=1)
  yone<-eig.vecs[2,1]/eig.vecs[1,1]
  xone<-xbar[2]-yone*xbar[1]
  abline(xone,yone,lty=2)
  ytwo<-eig.vecs[2,2]/eig.vecs[1,2]
  xtwo<-xbar[2]-ytwo*xbar[1]
  abline(xtwo,ytwo,lty=2)
  text(150,25,"yhat1")
  text(275,19,"yhat2")
  
  list("eigen vectors"=eig.vecs,"variance"=vars,"cumulative variance"=cum.var,"y hat"=yhat)
}

example_4_12(data)
```
```{r}
cor(yhat[,1],data[,1:2])
```
